An Experimental Analysis of the Influence of the Precision of Parameters in an Artificial Neural Network

### 2025-11-17 paper/main.tex
weiter

\section{Artificial Neural Networks}

A typical neuron in a neural network performs a weighted sum of its inputs,
followed by an activation function which results in the neuron output $h$
depicted in~\eqref{eq:activation}. Here, $\omega\in\mathbb{R}^n$ is the weight
vector, $b \in \mathbb{R}$ is the bias, $x \in \mathbb{R}^n$ is the input, and
$\sigma$ is a nonlinear activation function (e.g., sigmoid or ReLU).

@@

with $W$ and $b$ the weight matrix and bias of the last linear layer, $h$ the
layer input, and $e_y$ the one-hot vector of the target class. Here,
$\mathbbm{1}\{k=y\}$ is the indicator: it equals $1$ if $k=y$ and $0$
otherwise.

Memory and computational requirements during training are substantial, as
weights, gradients, activations, and parameters must be stored and updated
iteratively. Training and inference can both be optimized using precision
reduction and related techniques, but inference is more well suited to
optimizations because it only requires a forward pass.

\section{Memory Efficiency and Numerical Formats}

To reduce memory usage and computational load, precision-aware techniques such
as quantization and mixed-precision training are increasingly employed. Instead
of using 32-bit or 64-bit floating-point formats, parameters can be stored in
16-bit or even 8-bit fixed-point representations.
Equation~\ref{eq:floatingPointFormat} depicts the general floating point format
and terminology of significand, base, and exponent.
Tables~\ref{tab:binary-bits} and~\ref{tab:binary-precision} provide an overview
of the different floating point types and their respective characteristics.

\begin{align}\label{eq:floatingPointFormat}
	\frac{2469}{200}=12.345=\underbrace{12345}_{\text{significand}}\times\underbrace{10^{\overbrace{-3}^{\text{exponent}}}}_{\text{base}}
\end{align}

>>>>

\section{Artificial Neural Networks}
A typical neuron in a neural network performs a weighted sum of its inputs,
followed by an activation function which results in the neuron output $h$
depicted in~\eqref{eq:activation}. Here, $\omega\in\mathbb{R}^n$ is the weight
vector, $b \in \mathbb{R}$ is the bias, $x \in \mathbb{R}^n$ is the input, and
$\sigma$ is a nonlinear activation function (e.g., sigmoid or ReLU).

@@

with $W$ and $b$ the weight matrix and bias of the last linear layer, $h$ the
layer input, and $e_y$ the one-hot vector of the target class. Here,
$\mathbbm{1}\{k=y\}$ is the indicator: it equals $1$ if $k=y$ and $0$
otherwise.\\

\noindent
Memory and computational requirements during training are substantial, as
weights, gradients, activations, and parameters must be stored and updated
iteratively. Training and inference can both be optimized using precision
reduction and related techniques, but inference is more well suited to
optimizations because it only requires a forward pass.

\section{Memory Efficiency and Numerical Formats}
To reduce memory usage and computational load, precision-aware techniques such
as quantization and mixed-precision training are increasingly employed. Instead
of using 32-bit or 64-bit floating-point formats, parameters can be stored in
16-bit or even 8-bit fixed-point representations.
Equation~\ref{eq:floatingPointFormat} depicts the general floating point format
and terminology of significand, base, and exponent.

\begin{align}\label{eq:floatingPointFormat}
	\frac{2469}{200}=12.345=\underbrace{12345}_{\text{significand}}\times{{\underbrace{10}_{\text{base}}}^{\overbrace{-3}^{\text{exponent}}}}
\end{align}

\noindent
Tables~\ref{tab:binary-bits} and~\ref{tab:binary-precision} provide an overview
of the different floating point types and their respective characteristics. The
\emph{sign}, \emph{exponent}, and \emph{significand} fields specify how many
bits are allocated to encode the sign of the number, its scale, and its
significant digits, respectively. The \emph{exponent bias} is a constant added
to the true exponent so that it can be stored as an unsigned integer. The
\emph{precision} (in bits) and corresponding \emph{decimal digits} tell how
many significant digits can be represented accurately in each format.
